Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes

### 2024-12-05 ms/ms.Rmd
ms: title and minor edits

title: Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes

@@

  Our models improved upon Norway's existing peat depth maps, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements. 

>>>>

title: Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes

@@

  Our models improved upon the national peat depth map, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements. 

### 2024-12-05 ms/ms.tex
ms: title and minor edits

\title{Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes}

@@

\begin{abstract}
Peatlands are Earth's most carbon-dense terrestrial ecosystems and crucial stores of soil carbon.
Their carbon density results largely from their depth, which can exceed ten meters.
However, peat depth is poorly mapped at scales relevant for land management and carbon accounting.
We evaluated whether digital soil mapping using remotely sensed data could improve existing maps of peat depth in Norway.
Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
We measured peat depth at 372 and 1878 locations using manual probing and ground-penetrating radar, then trained Random Forest models to predict depth at 10 m resolution from various combinations of predictors.
The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
Terrain variables were more important predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
Adding radiometric predictors improved model performance at one site but degraded it at the other.
Our models improved upon Norway's existing peat depth maps, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.
\end{abstract}

@@

\caption{\label{tab:preds}Candidate predictors of peat depth.}
\centering
\begin{tabular}[t]{lll}
\hline
Group & Name & Description\\
\hline
radiometric & radK & Potassium ground concentration\\
 & radTh & Thorium ground concentration\\
 & radU & Uranium ground concentration\\
 & radTC & Total count of gamma radiation\\
terrain & elevation & Mean elevation\\
 & slope1m & Mean of 1 m slope\\
 & TPI1m & Mean of 1 m topographic position index\\
 & TRI1m & Mean of 1 m terrain ruggedness index\\
 & roughness1m & Mean of 1 m roughness\\
 & slope10m & 10 m slope\\
 & TPI10m & 10 m topographic position index\\
 & TRI10m & 10 m terrain ruggedness index\\
 & roughness10m & 10 m roughness\\
 & MRVBF & Multi-resolution valley bottom flatness\\
 & TWI5m & Mean of 5 m topographic wetness index\\
 & TWI10m & 10 m topographic wetness index\\
 & TWI20m & Bilinear interpolation of 20 m topographic wetness index\\
 & TWI50m & Bilinear interpolation of 50 m topographic wetness index\\
 & DTW2500 & Depth-to-water index, flow initiation area of 0.25 ha\\
 & DTW5000 & Depth-to-water index, flow initiation area of 0.5 ha\\
 & DTW10000 & Depth-to-water index, flow initiation area of 1 ha\\
 & DTW20000 & Depth-to-water index, flow initiation area of 2 ha\\
 & DTW40000 & Depth-to-water index, flow initiation area of 4 ha\\
 & DTW80000 & Depth-to-water index, flow initiation area of 8 ha\\
 & DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
existing map & DMK & DMK peat depth class, categorical with 3 levels\\
\hline
\end{tabular}
\end{table}

>>>>

\title{Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes}

@@

\begin{abstract}
Peatlands are Earth's most carbon-dense terrestrial ecosystems and crucial stores of soil carbon.
Their carbon density results largely from their depth, which can exceed ten meters.
However, peat depth is poorly mapped at scales relevant for land management and carbon accounting.
We evaluated whether digital soil mapping using remotely sensed data could improve existing maps of peat depth in Norway.
Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
We measured peat depth at 372 and 1878 locations using manual probing and ground-penetrating radar, then trained Random Forest models to predict depth at 10 m resolution from various combinations of predictors.
The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
Terrain variables were more important predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
Adding radiometric predictors improved model performance at one site but degraded it at the other.
Our models improved upon the national peat depth map, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.
\end{abstract}

@@

\caption{\label{tab:preds}Candidate predictors of peat depth.}
\centering
\begin{tabular}[t]{lll}
\hline
Group & Name & Description\\
\hline
radiometric & radK & Potassium ground concentration\\
 & radTh & Thorium ground concentration\\
 & radU & Uranium ground concentration\\
 & radTC & Total count of gamma radiation\\
terrain & elevation & Mean elevation\\
 & slope1m & Mean of 1 m slope\\
 & TPI1m & Mean of 1 m topographic position index\\
 & TRI1m & Mean of 1 m terrain ruggedness index\\
 & roughness1m & Mean of 1 m roughness\\
 & slope10m & 10 m slope\\
 & TPI10m & 10 m topographic position index\\
 & TRI10m & 10 m terrain ruggedness index\\
 & roughness10m & 10 m roughness\\
 & MRVBF & Multi-resolution valley bottom flatness\\
 & TWI5m & Mean of 5 m topographic wetness index\\
 & TWI10m & 10 m topographic wetness index\\
 & TWI20m & Bilinear interpolation of 20 m topographic wetness index\\
 & TWI50m & Bilinear interpolation of 50 m topographic wetness index\\
 & DTW2500 & Depth-to-water index, flow initiation area of 0.25 ha\\
 & DTW5000 & Depth-to-water index, flow initiation area of 0.5 ha\\
 & DTW10000 & Depth-to-water index, flow initiation area of 1 ha\\
 & DTW20000 & Depth-to-water index, flow initiation area of 2 ha\\
 & DTW40000 & Depth-to-water index, flow initiation area of 4 ha\\
 & DTW80000 & Depth-to-water index, flow initiation area of 8 ha\\
 & DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
national map & DMK & DMK peat depth class, categorical with 3 levels\\
\hline
\end{tabular}
\end{table}
